---
title: "Abundance Graphs"
format: html
---

# Run intro scripts

```{r}
# source("scripts/01_metadata_import_clean.R") # Run only if changes made to metadata
# source("scripts/02_import_mag_sylph_16s.R")

load("output/data/mt_16s.RData")
load("output/data/mt_mag.RData")
load("output/data/mt_sylph.RData")
```

# Load Required Libraries

```{r}
library(plotly)
library(ggalluvial)
library(ggarrow)
library(tidyverse)
library(microeco)
library(ggrepel)
```

# Final plots (atm)

```{r}

# 16S, Robust Aitchison, NMDS and stressplot

  tb2 <- trans_beta$new(mt_16s, group = "strat_season_2", measure = "bray")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

stress <- vegan::stressplot(tb2$res_ordination$model)
stress

plot <- tb2$plot_ordination(plot_color = "strat_season_2", plot_type = c("point", "ellipse"), NMDS_stress_pos = NULL, ellipse_level = 0.95, add_sample_label = "sample") +
  geom_arrow_chain(colour = "black") + 
    ggtitle("NMDS Ordination of WM Data", subtitle = paste("Stress = ", round(tb2$res_ordination$model$stress, 2))) + 
    geom_label_repel(label = mt_16s$sample_table$sample, color = "black") +
    labs(color = "Season") +
    guides(fill = "none") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 3)
    )
  plot
```

# Tests

```{r}
# 16S and MAGS PCoA and NMDS with a bunch of different measures. To generate all plots to compare, run this whole block. 

measures <- c("robust.aitchison", "bray", "aitchison")
methods <- c("NMDS", "PCoA")
data <- c("mt_mag","mt_16s")

for(i in 1:length(methods)) {
  for(ii in 1:length(measures)) {
tb2 <- trans_beta$new(mt_16s, group = "strat_season", measure = measures[ii])
tb2$cal_ordination(method = methods[i])
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

title <- paste("mt_16s", methods[i], measures[ii])
plot <- tb2$plot_ordination(plot_color = "strat_season", plot_type = c("point", "ellipse"), ellipse_level = 0.95, loading_arrow = TRUE, loading_text_italic = TRUE) +
  geom_arrow_chain(colour = "black") + 
    ggtitle(title) + 
    geom_label_repel(label = mt_16s$sample_table$sample, color = "black") +
    labs(color = "Season") +
    guides(fill = "none") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 3)
    )
print(plot)
path <- paste0("output/plots/beta_div/", title, ".png")
ggsave(path, plot)
  }
}

# MAGs R aitchison, PCoA

measures <- c("robust.aitchison", "bray", "aitchison", "wei_unifrac")
methods <- c("NMDS", "PCoA")
data <- c("mt_mag","mt_16s")

for(i in 1:length(methods)) {
  for(ii in 1:length(measures)) {
tb2 <- trans_beta$new(mt_mag, group = "strat_season", measure = measures[ii])
tb2$cal_ordination(method = methods[i])
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

title <- paste("mt_mag", methods[i], measures[ii])
plot <- tb2$plot_ordination(plot_color = "strat_season", plot_type = c("point", "ellipse"), ellipse_level = 0.95, loading_arrow = TRUE, loading_text_italic = TRUE) +
  geom_arrow_chain(colour = "black") + 
    ggtitle(title) + 
    geom_label_repel(label = mt_16s$sample_table$sample, color = "black") +
    labs(color = "Season") +
    guides(fill = "none") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 3)
    )
print(plot)
path <- paste0("output/plots/beta_div/", title, ".png")
ggsave(path, plot)
  }
}
```

## Abundance Plots

```{r}

# Set rank (k, p, c, etc) and number of taxa to include
rank <- "c"
n <- 20

# Create abundance object
ts1 <- trans_abund$new(mt_mag, taxrank = rank, ntaxa = n)
ts1$sample_table$strat_season <- factor(ts1$sample_table$strat_season, 
  levels = c("summer", "fall", "winter", "spring"), 
  labels = c("Summer", "Fall", "Winter", "Spring"),
   ordered = T)
# Plot abunances
abund_s <- ts1$plot_bar(others_color = "grey70", xtext_keep = TRUE, legend_text_italic = FALSE, facet = "strat_season") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size=9)) +
  ggtitle("MAGs Data", subtitle = paste(n, "most abundant", rank))
abund_s

plotly_build(abund_s)
mt_16s$sample_table$strat_season <- factor(mt_16s$sample_table$strat_season, 
  levels = c("summer", "fall", "winter", "spring"), 
  labels = c("Summer", "Fall", "Winter", "Spring"),
   ordered = T)
# Do the same but for 16s data

rank <- "g"
n <- 30
ts1 <- trans_abund$new(mt_16s, taxrank = rank, ntaxa = n)
ts1$sample_table$strat_season <- factor(ts1$sample_table$strat_season, 
  levels = c("summer", "fall", "winter", "spring"), 
  labels = c("Summer", "Fall", "Winter", "Spring"),
   ordered = T)
# Plot abunances
abund_16 <- ts1$plot_bar(others_color = "grey70", xtext_keep = TRUE, legend_text_italic = FALSE, facet = "strat_season") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size=9)) +
  ggtitle("16s Data", subtitle = paste(n, "most abundant", rank))
abund_16

plotly_build(abund_16)

subset <- clone(mt_16s) 

subset$tax_table <- subset$tax_table|> 
  filter(c == "c__Bacteroidia")

subset$tidy_dataset()$cal_abund()

t1 <- trans_abund$new(dataset = subset, taxrank = "g", ntaxa = n, show = 0, high_level = "o", high_level_fix_nsub = 10)
t1$plot_bar(ggnested = T)
```

## Plot single taxa abundance

# Plot just a single taxa abundance

```{r}
# Clone dataset
taxa <- "Planctomycetia" # taxa of interest
n <- 20 # Number of taxa to included in final graph
mt_sylph_subset <- clone(mt_sylph)

###mt_sylph_subset include the taxa below
mt_sylph_subset$tax_table <- mt_sylph_subset$tax_table |> 
  filter(c == taxa)
mt_sylph_subset$tidy_dataset()

# check
mt_sylph_subset # OK
head(mt_sylph_subset$tax_table)

# recalculate abundances (needed?)
mt_sylph_subset$cal_abund()
mt_sylph_subset$taxa_abund

# plot:

# t1 <- trans_abund$new(dataset = mt_sylph_subset, taxrank = "f", ntaxa = n)
# t1$plot_bar() |> ggplotly()

# Nested plot
t1 <- trans_abund$new(dataset = mt_sylph_subset, taxrank = "f", ntaxa = n, show = 0, high_level = "o", high_level_fix_nsub = 10)
t1$plot_bar(ggnested = T) |> ggplotly()
```

# Single taxa abundance based on top 20 families

```{r}
rank = "o"
ntaxa = 30
# Get vector of 20 most abundant families
t1 <- trans_abund$new(dataset = mt_mag, taxrank = rank, ntaxa = ntaxa)
t1$plot_bar(facet = "strat_season") +
     ggtitle("MAGs", subtitle = paste("top", ntaxa, rank))
taxa_names <- t1$data_taxanames
taxa_names

# Get vector of 20 most abundant families
t1 <- trans_abund$new(dataset = mt_16s, taxrank = rank, ntaxa = ntaxa)
t1$plot_bar(facet = "strat_season") +
     ggtitle("16s", subtitle = paste("top", ntaxa, rank))
taxa_names <- t1$data_taxanames
taxa_names

for (i in 1:length(taxa_names)) {
    subset <- clone(mt_16s)
    #subset include the taxa below
    subset$tax_table <- subset$tax_table |> 
    filter(f == paste0(rank, "__",taxa_names[i]))
    # filter(f == taxa_names[i])
    subset$tidy_dataset()

    # recalculate abundances (needed?)
    subset$cal_abund()
    subset$taxa_abund

    # plot:

    t1 <- trans_abund$new(dataset = subset, taxrank = "g", ntaxa = 20)
    plot <- t1$plot_bar(facet = "strat_season") +
    ggtitle(taxa_names[i], subtitle = "16s")
    print(plot) 
        }

    # Get vector of 20 most abundant families
t1 <- trans_abund$new(dataset = mt_sylph, taxrank = rank, ntaxa = ntaxa)
taxa_names <- t1$data_taxanames
taxa_names
t1$plot_bar(facet = "strat_season") +
  ggtitle("MAGs", subtitle = paste("top", ntaxa, rank))

for (i in 1:length(taxa_names)) {
    subset <- clone(mt_sylph)
    #subset include the taxa below
    subset$tax_table <- subset$tax_table |> 
    filter(f == taxa_names[i])
    subset$tidy_dataset()

    # recalculate abundances (needed?)
    subset$cal_abund()
    subset$taxa_abund

    # plot:

    t1 <- trans_abund$new(dataset = subset, taxrank = "g", ntaxa = 20)
    plot <- t1$plot_bar(facet = "strat_season") +
    ggtitle(taxa_names[i], subtitle = "MAG")
    print(plot) 
        }
```

```{r}
# Miscellaneous tests

abundant_taxa(mt_16s, 20, rank = "f", plot = T)
abundant_taxa <- function(meco_table, ntaxa = NULL, rank = "f", plot = F) {
    t1 <- trans_abund$new(dataset = meco_table, taxrank = rank, ntaxa = ntaxa)
    taxa_names <- t1$data_taxanames
    if(!plot) {
        next
    } else {
        for (i in 1:length(taxa_names)) {
            subset <- clone(meco_table)

    #subset include the taxa below
    subset$tax_table <- subset$tax_table |> 
    filter(o == taxa_names[i])
    subset$tidy_dataset()

    # recalculate abundances (needed?)
    mt_sylph_subset$cal_abund()
    mt_sylph_subset$taxa_abund

    # plot:

    t1 <- trans_abund$new(dataset = subset, taxrank = "g", ntaxa = 20)
    t1$plot_bar(facet = "strat_season") +
    ggtitle(taxa_name[i])
            }
        }
    return(taxa_names)
    }


taxa <- abundant_taxa(mt_sylph, 20, "f")
```

```{r}
# Clone dataset
taxa <- "Rhodoglobus" # taxa of interest
n <- 5 # Number of taxa to included in final graph
mt_sylph_subset <- clone(mt_sylph)

###mt_sylph_subset include the taxa below
mt_sylph_subset$tax_table <- mt_sylph_subset$tax_table |> 
  filter(g == taxa)
mt_sylph_subset$tidy_dataset()

# check
mt_sylph_subset # OK
head(mt_sylph_subset$tax_table)

# recalculate abundances (needed?)
mt_sylph_subset$cal_abund()
mt_sylph_subset$taxa_abund

# plot:

t1 <- trans_abund$new(dataset = mt_sylph_subset, taxrank = "s", ntaxa = n)
t1$plot_bar(facet = "strat_season") +
  ggtitle(taxa)
```

```{r}
# Plot alluvial

ts1$plot_bar(bar_full = F, use_alluvium = T, others_color = "grey70", xtext_keep = TRUE, legend_text_italic = FALSE, clustering = T) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size=9))

# Double Abudances plot
  # Sylph data
ts2 <- trans_abund$new(dataset = mt_sylph, taxrank = "f", ntaxa = 20)
#  show = 0, high_level = "p", high_level_fix_nsub = 2)

ts2$plot_bar(ggnested = F, xtext_angle = 45) |> 
    ggplotly()
    # 16S data
  t2 <- trans_abund$new(dataset = mt_16s, taxrank = "f", ntaxa = 20, show = 0, high_level = "p", high_level_fix_nsub = 6)

t2$plot_bar(ggnested = T, xtext_angle = 45) +
    ggtitle("16S") |> 
    ggplotly()

# Heat Map

t3 <- trans_abund$new(dataset = mt, taxrank = "g", ntaxa = 40)
g1 <- t3$plot_heatmap(xtext_keep = T, withmargin = FALSE, plot_breaks = c(0.01, 0.1, 1, 10))
g1
g1 + theme(axis.text.y = element_text(face = 'italic'), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size=9))
```

# Alpha Diversity

```{r}

mt_sylph$alpha_diversity |>
  rownames_to_column("sample") |> 
  ggplot(aes(sample, Simpson, group = 1)) +
  geom_point() +
  geom_line()
```

# Beta Diversity

```{r}
#Add in quantiles?
mt_sylph$sample_table <- mt_sylph$sample_table |> 
  mutate(temp_quant = cut_number(temp_c_24h_med, n = 4))
tb2 <- trans_beta$new(mt_sylph, group = "strat_season", measure = "bray")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

tb2$plot_ordination(plot_color = "strat_season", plot_shape = "strat_season", plot_type = c("point", "ellipse"), NMDS_stress_text_prefix = "", ellipse_level = 0.95) +
  #   geom_text(
  #   data = tb2$dataset$otu_table, # Filter data for the label layer
  #   aes(label = sample),
  #   vjust = -1.1, # Adjust vertical position
  #   hjust = 0.5,  # Adjust horizontal position
  #   color = "black"
  # ) +
  # geom_arrow_chain(colour = "black") + 
    ggtitle("Sylph NMDS") +
  # scale_color_viridis_b() +
  NULL

tb2 <- trans_beta$new(mt_16s, group = "strat_season", measure = "robust.aitchison")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

tb2$plot_ordination(plot_color = "strat_season", plot_shape = "strat_season", NMDS_stress_text_prefix = "",plot_type = c("point", "ellipse"), ellipse_level = 0.95) +
  geom_arrow_chain(colour = "black") + 
    ggtitle("16s NMDS") +
  #  scale_color_viridis_b() +
  NULL

vegan::stressplot(tb2$res_ordination$model)
  # Microeco PCoA

tb2 <- trans_beta$new(mt_16s, group = "strat_season", measure = "robust.aitchison")
tb2$cal_ordination(method = "PCoA")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

test <- tb2$plot_ordination(plot_color = "strat_season", plot_shape = "strat_season", plot_type = c("point", "ellipse"), ellipse_level = 0.95) +
  #   geom_text(
  #   data = tb2$dataset$otu_table, # Filter data for the label layer
  #   aes(label = colnames(tb2$dataset$otu_table)),
  #   vjust = -1.1, # Adjust vertical position
  #   hjust = 0.5,  # Adjust horizontal position
  #   color = "black"
  # ) +
  geom_arrow_chain(colour = "black") + 
    ggtitle(paste0("aitchison-  MAGs PCoA")) +
  # scale_color_viridis_b() +
  NULL
test

  tb2 <- trans_beta$new(mt_mag, group = "strat_season", measure = "bray")
tb2$cal_ordination(method = "PCoA")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

test <- tb2$plot_ordination(plot_color = "strat_season", plot_shape = "strat_season", plot_type = c("point", "ellipse"), ellipse_level = 0.95) +
  #   geom_text(
  #   data = tb2$dataset$otu_table, # Filter data for the label layer
  #   aes(label = colnames(tb2$dataset$otu_table)),
  #   vjust = -1.1, # Adjust vertical position
  #   hjust = 0.5,  # Adjust horizontal position
  #   color = "black"
  # ) +
  geom_arrow_chain(colour = "black") + 
    ggtitle(paste0("bray-  MAGs PCoA")) +
  # scale_color_viridis_b() +
  NULL
test

  tb2 <- trans_beta$new(mt_sylph, group = "strat_season", measure = "bray")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

test <- tb2$plot_ordination(plot_color = "strat_season", plot_shape = "strat_season", plot_type = c("point", "ellipse"), ellipse_level = 0.95) +
  #   geom_text(
  #   data = tb2$dataset$otu_table, # Filter data for the label layer
  #   aes(label = colnames(tb2$dataset$otu_table)),
  #   vjust = -1.1, # Adjust vertical position
  #   hjust = 0.5,  # Adjust horizontal position
  #   color = "black"
  # ) +
  geom_arrow_chain(colour = "black") + 
    ggtitle(paste0("bray- sylphs NMDS")) +
  # scale_color_viridis_b() +
  NULL
test


  tb2 <- trans_beta$new(mt_mag, group = "strat_season", measure = "bray")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

test <- tb2$plot_ordination(plot_color = "strat_season", plot_shape = "strat_season", plot_type = c("point", "ellipse"), ellipse_level = 0.95) +
  geom_arrow_chain(colour = "black") + 
      geom_text(
    aes(label = tb2$sample_table$sample),
    vjust = -1.1, # Adjust vertical position
    hjust = 0.5,  # Adjust horizontal position
    color = "black"
  ) +
    ggtitle(paste0("bray-  16s NMDS")) +
  # scale_color_viridis_b() +
  NULL
test

  tb2 <- trans_beta$new(mt_mag, group = "strat_season", measure = "unwei_unifrac")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)
```

## Beta Diversity but with temperature quantile

```{r}

#Add in quantiles?
mt_sylph$sample_table <- mt_sylph$sample_table |> 
  mutate(temp_quant = cut_number(temp_c_24h_med, n = 4))
tb2 <- trans_beta$new(mt_sylph, group = "temp_quant", measure = "bray")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

tb2$plot_ordination(plot_color = "strat_season", plot_shape = "temp_quant", plot_type = c("point", "ellipse"), NMDS_stress_text_prefix = "", ellipse_level = 0.95) +
  #   geom_text(
  #   data = tb2$dataset$otu_table, # Filter data for the label layer
  #   aes(label = sample),
  #   vjust = -1.1, # Adjust vertical position
  #   hjust = 0.5,  # Adjust horizontal position
  #   color = "black"
  # ) +
  # geom_arrow_chain(colour = "black") + 
    ggtitle("Sylph NMDS") +
  # scale_color_viridis_b() +
  NULL
```

Try the same thing with 16s data

```{r}

#Add in quantiles?
mt_16s$sample_table <- mt_16s$sample_table |> 
  mutate(temp_quant = cut_number(temp_c_24h_med, n = 4))
tb2 <- trans_beta$new(mt_16s, group = "temp_quant", measure = "robust.aitchison")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

tb2$plot_ordination(plot_color = "strat_season", plot_shape = "temp_quant", plot_type = c("point", "ellipse"), NMDS_stress_text_prefix = "", ellipse_level = 0.95) +
  #   geom_text(
  #   data = tb2$dataset$otu_table, # Filter data for the label layer
  #   aes(label = sample),
  #   vjust = -1.1, # Adjust vertical position
  #   hjust = 0.5,  # Adjust horizontal position
  #   color = "black"
  # ) +
  # geom_arrow_chain(colour = "black") + 
    ggtitle("Sylph NMDS") +
  # scale_color_viridis_b() +
  NULL
```

## 16s Beta Diversity

```{r}

tb2 <- trans_beta$new(mt_16s, group = "strat_season", measure = "robust.aitchison")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

plot <- tb2$plot_ordination(plot_color = "strat_season", NMDS_stress_text_prefix = "Stress: ", plot_type = c("point", "ellipse"), ellipse_level = 0.95, add_sample_label = "sample") +
  # geom_arrow_chain(colour = "black") + 
        ggtitle("NMDS Ordination of WM Data", subtitle = paste("Stress = ", round(tb2$res_ordination$model$stress, 2))) + 
    # geom_label_repel(label = mt_mag$sample_table$sample, color = "black") +
    labs(color = "strat_season") +
    guides(fill = "none") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 3)
    )
  plot

plot + labs(color = "Season", shape = "test")
  class(plot)

  #  scale_color_viridis_b() 

  tb2 <- trans_beta$new(mt_16s, group = "strat_season", measure = "bray")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

plot <- tb2$plot_ordination(plot_color = "strat_season", NMDS_stress_text_prefix = "Stress: ", plot_type = c("point", "ellipse"), ellipse_level = 0.95, add_sample_label = "sample") +
  geom_arrow_chain(colour = "black") + 
    ggtitle("16s NMDS-bray")

  plot

  tb2 <- trans_beta$new(mt_16s, group = "strat_season", measure = "aitchison")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

plot <- tb2$plot_ordination(plot_color = "strat_season", NMDS_stress_text_prefix = "Stress: ", plot_type = c("point", "ellipse"), ellipse_level = 0.95, add_sample_label = "sample") +
  geom_arrow_chain(colour = "black") + 
    ggtitle("16s NMDS - aitchison")
  plot

  tb2 <- trans_beta$new(mt_16s, group = "strat_season", measure = "robust.aitchison")
tb2$cal_ordination(method = "NMDS")
tb2$cal_manova(manova_all = TRUE)
print(tb2$res_manova)

plot <- tb2$plot_ordination(plot_color = "strat_season", plot_type = c("point", "ellipse"), NMDS_stress_pos = NULL, ellipse_level = 0.95) +
  geom_arrow_chain(colour = "black") + 
    ggtitle("NMDS Ordination of WM Data", subtitle = paste("Stress = ", round(tb2$res_ordination$model$stress, 2))) + 
    geom_label_repel(label = mt_16s$sample_table$sample, color = "black") +
    labs(color = "Season") +
    guides(fill = "none") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 3)
    )
  plot



plot <- tb2$plot_ordination(plot_color = "strat_season", plot_type = c("point", "ellipse"), NMDS_stress_pos = NULL, ellipse_level = 0.95) +
  # geom_arrow_chain(colour = "black") + 
    ggtitle("NMDS Ordination of WM Data", subtitle = paste("Stress = ", round(tb2$res_ordination$model$stress, 2))) + 
    # geom_label_repel(label = mt_mag$sample_table$sample, color = "black") +
    labs(color = "Season") +
    guides(fill = "none") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 3)
    )
  plot

  plot <- tb2$plot_ordination(plot_type = "point", NMDS_stress_pos = NULL, ellipse_level = 0.95) +
  # geom_arrow_chain(colour = "black") + 
    ggtitle("NMDS Ordination of WM Data", subtitle = paste("Stress = ", round(tb2$res_ordination$model$stress, 2))) + 
    # geom_label_repel(label = mt_mag$sample_table$sample, color = "black") +
    labs(color = "Season") +
    guides(fill = "none") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 3)
    )
  plot
stressplot(tb2$res_ordination$model)
screeplot(tb2$res_ordination$model)

```

```{r}

  read_tsv
  nrow(read_tsv("data/gtdb_bin_tax/gtdbtk.ar53.summary.tsv"))
```

```{r}
# Test to see how it works to pull in the microtrait data here

load("output/data/mt_mag_large.RData")
library(microeco)
library(tidyverse)
library(ggpattern)
# mt_mag_large$cal_abund()


# Plot abundances

# subset$tax_table <- subset$tax_table|> 
#   filter(c == "c__Bacteroidia")

# subset$tidy_dataset()$cal_abund()

# t1 <- trans_abund$new(dataset = subset, taxrank = "g", ntaxa = n, show = 0, high_level = "o", high_level_fix_nsub = 10)
# t1$plot_bar(ggnested = T)

#Import the column names so I can take a look at them

microtrait_cols <- read_csv("output/data/microtrait_columns.csv")


trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = "poss_heterotroph")$plot_bar(ggnested = TRUE) +
  ggtitle("50 most abundant families, split by possible heterotroph based on microtrait (T/F)")

trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = "poss_chemotroph")$plot_bar(ggnested = TRUE) +
  ggtitle("50 most abundant families, split by possible chemotroph based on microtrait (T/F)")

trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = "poss_phototroph")$plot_bar(ggnested = TRUE) +
  ggtitle("50 most abundant families, split by possible phototroph based on microtrait (T/F)")


trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = "g1_resource_acquisition-substrate_degradation-complex_carbohydrate_depolymerization")$plot_bar(ggnested = TRUE) +
  ggtitle("50 most abundant families, split by possible carbohydrate depolymerization based on microtrait (T/F)")

param <- "g2_resource_use__chemotrophy__chemolithoheterotrophy__anaerobic_respiration__n_denitrification"
print(param)
trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = param)$plot_bar(ggnested = TRUE) +
  ggtitle(paste("50 most abundant families split by", param, "based on microtrait (T/F)"))

# Hmmm okay those are all true that's not super helpful, lets see what we can do with nitrate next

nitr_cols <- microtrait_cols |> 
  filter(str_detect(columns, "g.*nitr"))

for(i in 1:length(nitr_cols$columns)) {
  param <- nitr_cols$columns[i]
  
  title <- str_remove(param, "^.*__")
print(param)
plot <- trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = param)$plot_bar(ggnested = TRUE) +
  theme(legend.position = "none") +
  ggtitle(paste(title))
print(plot)
}

param <- "g3_resource_acquisition__substrate_assimilation__n_compounds__assimilatory_nitrate_reduction"

trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = param)$plot_bar(ggnested = TRUE) +
  # theme(legend.position = "none") +
  ggtitle(paste("50 families split by", param, "from microtrait T/F"))

mt_nitr <- clone(mt_mag_large) 

mt_nitr$tax_table <- mt_nitr$tax_table |> 
  mutate(n_trans = ifelse(g3_resource_acquisition__substrate_uptake__n_compound_transport__nitrate_transport == TRUE | g3_resource_acquisition__substrate_uptake__n_compound_transport__nitrite_transport == TRUE, TRUE, FALSE))


trans_abund$new(mt_nitr, taxrank = "f", ntaxa = 50, high_level = "n_trans")$plot_bar(ggnested = T) +
  ggtitle(paste("50 most abundant families, split by nitrate/nitrite transport based on microtrait (T/F)"))

# mt_mag_large$tax_table <- mt_mag_large$tax_table |> 
#   mutate(stripes = ifelse(poss_heterotroph == TRUE, "stripe", "none"))
```

Okay now I want to look at some other chemical pathways

```{r}
# Let's continue on the nitrogen train and look at ammonia

phos_cols <- microtrait_cols |> 
  filter(str_detect(columns, "^g.*phos.*"))


for(i in 1:length(phos_cols$columns)) {
  param <- phos_cols$columns[i]
  
  title <- str_remove(param, "^.*__")
print(param)
plot <- trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = param)$plot_bar(ggnested = TRUE) +
  theme(legend.position = "none") +
  ggtitle(paste(title))
print(plot)
}
```

Try some phosphorous
```{r}
amm_cols <- microtrait_cols |> 
  filter(str_detect(columns, "^g.*amm.*"))


for(i in 1:length(amm_cols$columns)) {
  param <- amm_cols$columns[i]
  
  title <- str_remove(param, "^.*__")
print(param)
plot <- trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = param)$plot_bar(ggnested = TRUE) +
  theme(legend.position = "none") +
  ggtitle(paste(title))
print(plot)
}


  param <- "g3_resource_acquisition__substrate_uptake__organophosphorus_compound_transport"
  
print(param)
trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = param)$plot_bar(ggnested = TRUE) +
  ggtitle(paste("50 families split by", param, "T/F"))

mt_mag_large$tax_table |> 
  filter(str_detect(f, "LD1")) |> 
  select(d, p, c, o, f, g, s) |> 
  unique()
```

## Try it all
```{r}

#First try to turn all true/false columns into a factor 
mt_mag_large$tax_table <- mt_mag_large$tax_table |> 
mutate(across(where(is.factor), ~ factor(.x, levels = c(TRUE, FALSE))))
g_cols  <- microtrait_cols |> 
  filter(str_detect(columns, "^g.*$"))


for(i in 83:length(g_cols$columns)) {
  param <- g_cols$columns[i]
  
  title <- str_remove(param, "^.*__")
print(param)
plot <- trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 116, high_level = param, high_level_fix_nsub = 116)$plot_bar(ggnested = TRUE) +
  # theme(legend.position = "none") +
  #   geom_bar_pattern(
  #   stat = "identity",
  #   aes(pattern = param),
  #   pattern_fill = "black",     # Color of the pattern lines
  #   pattern_angle = 45,         # Tilt the lines
  #   pattern_density = 0.2,      # How packed the lines are
  #   color = "black"             # Border color of the bars
  # ) +
  # scale_pattern_manual(values = c("stripe", "none")) +
  ggtitle(paste(title))
print(plot)
print(i)
readline()
}



for(i in 1:100) {
  param <- g_cols$columns[i]
  
  title <- str_remove(param, "^.*__")
print(param)
plot <- trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50, high_level = param)$plot_bar(ggnested = TRUE) +
  theme(legend.position = "none") +
  ggtitle(paste(title))
print(plot)
print(i)
readline()
}
colors <- distinct_palette(n = 41, pa = "brewerPlus")
t2 <- mt_mag_large$taxa_abund$bin |> 
  mutate(fam = str_extract(rownames(mt_mag_large$taxa_abund$bin), "(?<=b__).*$")) |> 
  rownames_to_column("rows") |>
  select(-rows) |>
  rowwise() |> 
  mutate(mean = mean(c_across(starts_with("WM")))) |> 
  ungroup() |> 
  arrange(desc(mean)) |> 
  select(fam) |> 
  as_vector()
# View(t2)
t3 <- trans_abund$new(mt_mag_large, taxrank = "bin", ntaxa = 116)
t3$data_taxanames <- t2[1:40]
test <- t3$plot_bar()

taxa <- test$data$Taxonomy

gen_times <- data.frame(fam = str_remove(mt_mag_large$tax_table$bin, "b__"), gen_time = mt_mag_large$tax_table$mingentime)

test$data <- test$data |> 
  mutate(gentime = ifelse(Taxonomy %in% gen_times$fam, gen_times$gen_time, as.numeric(NA)))

view(test$data)

# From here try to plot the gentime

test +
  scale_fill_gradient(data = test$data$gentime)

plot_data <- test$data

cols <- c(6, 22, 23, 24, 25)

plot_data |> 
  ggplot(aes(y= log(Abundance), x = cond_24h_avg, color = Taxonomy)) +
  geom_point() +
  # scale_color_manual(values = colors) +
  facet_wrap(~mixing, scales = "free") +
  geom_smooth(method = "lm",  se = T, linetype = "dashed", color = "black") +
  # ggpubr::stat_regline_equation(aes(label = after_stat(rr.label)), label.x = 4, 
  #                       label.y = 30) +
  theme_classic()


plot_data |> 
  select(9:21) |> 
  pivot_longer(cols = everything(), names_to = "var", values_to = "val") |> 
  ggplot(aes(x = val)) + 
  geom_histogram() +
  facet_wrap(~var, scales = "free")
```

##UBA search

```{r}
uba_table <- clone(mt_mag_large)

uba_table$tax_table <- uba_table$tax_table |> 
  filter(f == "f__UBA2100")

View(uba_table$tax_table)

uba_table$tidy_dataset()
uba_table$cal_abund()


t1 <- trans_abund$new(uba_table, taxrank = "bin", ntaxa = 10)$plot_bar()
t1

# Adding a new column 

uba_table$tax_table <- uba_table$tax_table |> 
  mutate(abundant = ifelse(bin == "b__WM14_S9_032" | bin == "b__WM15_S10_035", "winter", "summer")) |> 
  mutate(abundant = ifelse(bin == "b__WM12_S7_050", "not_very", abundant))
trans_abund$new(uba_table, taxrank = "", ntaxa = 5, high_level = param, high_level_fix_nsub = 3)$plot_bar(ggnested = TRUE)

lil_tab <- uba_table$tax_table |> 
  select(abundant, !matches("(hmm_|rule_)")) |> 
  arrange(abundant)

lil_tab <- lil_tab |> 
  mutate(across(where(is.factor), as.logical))

test <- lil_tab |> 
  filter_out(bin == "b__WM12_S7_050") |> 
  group_by(abundant) |> 
  summarise(across(matches("^g[0-9]{1}.*"), ~sum(.x))) |> 
  ungroup()
map(test, ~sum(.))


long_test <- test |> 
  column_to_rownames("abundant") |> 
  t() |> 
  as.data.frame() |> 
  rownames_to_column("microtrait") |> 
  rowwise() |> 
  mutate(sum = sum(c_across(summer:winter))) |> 
  ungroup() |> 
  filter_out(sum == 0|sum == 4) |> 
  filter(winter > summer)

print(long_test$microtrait)
trans_abund$new(mt_mag_large, taxrank = "f", ntaxa = 50)$plot_bar()


load("output/data/mt_gene_mag.RData")
sulfur <- mt_gene_mag$tax_table |> 
  filter(bin == "WM15_S10_035")

winter <- c("WM14_S9_032", "WM15_S10_035")
summer<- c("WM12_S7_038", "WM14_S9_031")
odd <- "WM12_S7_050"

all <- c(winter, summer, odd)

for(i in 1:length(all)) {
  file <- paste0("output/data/kegg_output/",all[i], "_kegg.tsv")
  ii <- read_tsv(file) |> 
    mutate(bin = all[i])
  
  if(i == 1){
    full <- ii
  } else {
    full <- bind_rows(full, ii)
  }
}

full <- full |> 
  mutate(season = ifelse(bin %in% summer, "summer", "winter")) |> 
  mutate(season = ifelse(bin %in% odd, "odd", season))

summer_genes <- full |> 
  filter(season == "summer")

winter_genes <- full |> 
  filter(season == "winter") |> 
  filter_out(id %in% summer_genes$id) |> 
  write_tsv(file = "output/data/kegg_output/uba_2100_winteronly_genes.tsv")

sum_text <- readLines("summer_kegg_uba2100.txt", warn = FALSE) |> 
  str_subset("^[A-Z].*", negate = T) |> 
  as_tibble_col() |> 
  mutate(path_num = str_extract(value, "^[0-9]{5}"), 
         path_nam = str_extract(value, "(?<=^.{6}).*(?=.{4}$)"), 
         matches = str_extract(value, "(?<=\\()[0-9]{1,2}(?=\\))"))


winter_only_text <- readLines("winter_only_kegg.txt", warn = FALSE) |> 
  str_subset("^[A-Z].*", negate = T) |> 
  as_tibble_col() |> 
  mutate(path_num = str_extract(value, "^[0-9]{5}"), 
         path_nam = str_extract(value, "(?<=^.{6}).*(?=.{4}$)"), 
         matches = str_extract(value, "(?<=\\()[0-9]{1,2}(?=\\))"))
#Look at the pathways that are only present in the winter
winter_only_text |> 
  filter_out(path_num %in% sum_text$path_num) |> 
  select(path_nam) |> 
  print(n = 100)
#Look at the pathways present in both but have different numbers of matches
winter_only_text |> 
  filter(path_num %in% sum_text$path_num) |>
  select(path_nam, matches) |> 
  print(n = 100)

```



```{r}
view(mt_mag_large$otu_table)
view(mt_mag_large$tax_table)

mt_mag_sliced <- clone(mt_mag_large)

mt_mag_sliced$tax_table <- mt_mag_sliced$tax_table |> 
  select(matches("g1_"))

mt_mag_sliced$tidy_dataset()
mt_mag_sliced$cal_abund()
cols <- mt_mag_sliced$tax_table |> 
  select(matches("g1_")) |> 
  colnames() 
t1 <- trans_diff$new(dataset = mt_mag_sliced, method = "lefse", group = "strat_season_2", alpha = 0.05, taxa_level = 'all', lefse_subgroup = NULL)

view(t1$res_diff)
sig_taxa <- t1$res_diff |> 
    filter (P.adj <=0.05)

save(t1, file = "t1_mag_large_lefse.RData")

load("t1_mag_large_lefse.RData")
t1$plot_diff_bar(threshold = 4.5)
# we show 20 taxa with the highest LDA (log10)
t1$plot_diff_bar(use_number = 1:30, width = 0.8, add_sig = T)

mt_mag_large$tax_table |> 
  colnames()


```


```{r}
load("output/data/mt_mag_large.RData")

uba_small <- mt_mag_large$tax_table |> 
  dplyr::filter(f == "f__UBA2100")


library(microeco)
library(tidyverse)
```

