---
title: "differential abundances"
format: html
---


```{r}
# Run intro scripts
# source("scripts/01_metadata_import_clean.R") # Run only if changes made to metadata
# source("scripts/02_import_mag_sylph_16s.R")
    load("output/data/mt_16s.RData")
    load("output/data/mt_mag.RData")
    load("output/data/mt_sylph.RData")
    load("output/data/mt_gene_mag.RData")
    load("output/data/t1_gene_ancombc2.RData")
```

# Load Required Libraries

```{r}
library(plotly)
library(ggalluvial)
library(ggarrow)
library(tidyverse)
library(microeco)
library(tidytree)
# BiocManager::install("ggtree")
# BiocManager::install("microbiome")
library(ggtree)
```

# Create trans_diff object


```{r}
# View sample table 
# view(mt_mag$sample_table)
t1 <- trans_diff$new(dataset = mt_mag, method = "ancombc2", group = "strat_season", fix_formula = "strat_season", alpha = 0.05, taxa_level = 'bin')
view(t1$res_diff)
sig_taxa <- t1$res_diff |> 
    filter (P.adj <=0.05)
t2 <- clone(t1)
t2$res_diff$Factors <- t1$res_diff$Factors |> 
    str_replace("\\((i|I)ntercept\\)", "summer") |> 
    str_replace("strat_season.L", "fall") |> 
    str_replace("strat_season.Q", "winter") |> 
    str_replace("strat_season.C", "spring")
t2$res_diff <- t2$res_diff |> 
    filter(P.adj < 0.05) 
t2$plot_diff_bar(keep_full_name = TRUE, heatmap_cell = "P.adj", heatmap_sig = "Significance", heatmap_x = "Factors", heatmap_y = "Taxa")


fig_pair <- t2$res_diff |> 
    filter(t2$res_diff$P.adj < 0.05) |> 
    mutate(Taxa = str_remove(Taxa, "\\|WM.*")) |> 
    mutate(Taxa = str_remove(Taxa, "d__Bacteria\\|")) |> 
    # mutate(Taxa = str_remove(Taxa, "\\|s__.*")) |> 
    distinct(Taxa, .keep_all = T) |> 
    ggplot(aes(x = Factors, y = Taxa, fill = lfc)) + 
    geom_tile(color = "black") +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         na.value = "white",
                         name = NULL) +
    geom_text(aes(Factors, Taxa, label = lfc), size = 5, color = "black") +
    scale_color_identity(guide = FALSE) +
    labs(x = NULL, y = NULL, title = "MAG Log fold changes compared to summer, (p < 0.1)") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), text = element_text(color = "black", size = 15))
fig_pair





# we show 20 taxa with the highest LDA (log10)
t1$plot_diff_bar(use_number = 1:30, width = 0.8, add_sig = T)


```

## Interpretation
It looks like the UBA2100 family is super prevalent during the winter. 
Sediminibacterium is around a lot during the fall season?


# 16s Analysis

```{r}
t1 <- trans_diff$new(dataset = mt_16s, method = "ancombc2", group = "strat_season", fix_formula = "strat_season", alpha = 0.05, taxa_level = 'otu')
save(t1, file = "output/data/t1_16s_ancombc2.RData")
view(t1$res_diff)

t1$res_diff <- 
    t1$res_diff |> 
    separate_wider_delim(Taxa, delim = "|", names = c("bin", "l1", "l2", "l3", "l4", "func", "gene"))
sig_taxa <- t1$res_diff |> 
    filter (P.adj <=0.05)
t2 <- clone(t1)
t2$res_diff$Factors <- t1$res_diff$Factors |> 
    str_replace("\\((i|I)ntercept\\)", "summer") |> 
    str_replace("strat_season.L", "fall") |> 
    str_replace("strat_season.Q", "winter") |> 
    str_replace("strat_season.C", "spring")


fig_pair <- t2$res_diff |> 
    filter(t2$res_diff$P.adj < 0.05) |> 
    distinct(gene, .keep_all = T) |> 
    ggplot(aes(x = Factors, y = gene, fill = lfc)) + 
    geom_tile(color = "black") +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                         na.value = "white",
                         name = NULL) +
    geom_text(aes(Factors, gene, label = lfc), size = 5, color = "black") +
    scale_color_identity(guide = FALSE) +
    labs(x = NULL, y = NULL, title = "16S Log fold changes compared to summer, (p < 0.1)") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5), text = element_text(color = "black", size = 15))
fig_pair


t1 <- trans_diff$new(dataset = mt_16s, method = "ALDEx2_k", group = "strat_season", fix_formula = "strat_season", alpha = 0.05, taxa_level = 'otu')
save(t1, file = "output/data/t1_16s_aldex2.RData")

```
```{r}
# Quick plot of temperature at our site


```

# 16s abundances
```{r}

clone_16s <- clone(mt_16s)
clone_16s$filter_taxa(
  rel_abund = 0.00001,  # At least 0.001% in any sample
)
clone_16s$tidy_dataset()
t1 <- trans_diff$new(dataset = clone_16s, method = "ALDEx2_kw", group = "strat_season", alpha = 0.05, taxa_level = 'otu')
save(t1, file = "output/data/16s_alde.RData")
view(t1$res_diff)
t1$plot_diff_cladogram()
t1$plot_diff_bar(threshold = 4.5)
# we show 20 taxa with the highest LDA (log10)
t1$plot_diff_bar(use_number = 1:30, width = 0.8, add_sig = T)
t1 <- trans_diff$new(dataset = mt_mag, method = "lefse", group = "strat_season", alpha = 0.05, taxa_level = 's', lefse_subgroup = NULL)
view(t1$res_diff)

t1$plot_diff_bar(threshold = 4.5)
# we show 20 taxa with the highest LDA (log10)
t1$plot_diff_bar(use_number = 1:30, width = 0.8, add_sig = T)


t1 <- trans_diff$new(dataset = mt_16s, method = "lefse", alpha = 0.1, group = "strat_season", lefse_subgroup = '')

t1$plot_diff_bar(threshold = 4)
# we show 20 taxa with the highest LDA (log10)
t1$plot_diff_bar(use_number = 1:30, width = 0.8)

t1$plot_diff_abund(use_number = 1:20)
t1$plot_diff_abund(use_number = 1:20, fill = "strat_season", alpha = 0.05, add_sig = F)

t1$plot_diff_cladogram()


```

"LEfSe scores can be interpreted as the degree of consistent difference in relative abundance between features in the two classes of analyzed microbial communities." - Segata et al 2011

```{r}
# See if I can quick figure out collinear variables

vifs <- mt_mag$sample_table |> 
    select(where(is.numeric)) |> 
    select(-c(depth, mtemp_2_c_24h_avg, mtemp_c_24h_avg, temp_c_14d_avg)) |> 
    drop_na() |> 
    cor()

corrplot::corrplot(vifs)
table(mt_mag$sample_table$strat_season, mt_mag$sample_table$use_maestro)

```

# Trans Func exploration 


```{r}
t2 <- trans_func$new(mt_16s)
t2$for_what <- 'prok'

t2$cal_func(prok_database = "FAPROTAX")

network <- trans_network$new(dataset = mt_16s, cal_cor = "base", taxa_level = "OTU", filter_thres = 0.0001, cor_method = "spearman")
network$cal_network(p_thres = 0.01, COR_cut = 0.7)
network$cal_module()
# convert module info to microtable object
meco_module <- network$trans_comm(use_col = "module")
meco_module_func <- trans_func$new(meco_module)
meco_module_func$for_what <- "prok"
meco_module_func$cal_func(prok_database = "FAPROTAX")
meco_module_func$cal_func_FR(abundance_weighted = FALSE, perc = TRUE)
meco_module_func$plot_func_FR()
```