---
title: "05_environmental_data"
format: html
---

# Load Libraries

```{r}
library(microeco)
library(tidyverse)
library(skimr)
```

# Load Data

```{r}
load("output/data/mt_mag.RData")
load("output/data/mt_16s.RData")
load("output/data/mt_sylph.RData")
load("output/data/temp_df_long.RData")
```

---

# Environmental Analysis — MAGs

## Build trans_env Objects

Two objects are created: one with interpolated NAs filled (`complete_na = T`) for complete-case analysis, and one without (`complete_na = F`) to track which values were originally missing.

```{r}

cols <- ncol(mt_mag$sample_table)
# AIC: t1 fills NAs (interpolated), t2 keeps them to flag which values were missing
t1 <- trans_env$new(dataset = mt_mag, env_cols = c(1:cols), standardize = F, complete_na = T)
t2 <- trans_env$new(dataset = mt_mag, env_cols = c(1:cols), standardize = F, complete_na = F)

env_long_complete <- t1$data_env |> 
  select(-c(depth, sample_date, use_maestro, strat_season_2, solar_season, mixing, month)) |> 
  pivot_longer(cols = -c(date, strat_season, sample), names_to = "var", values_to = "value") |> 
  mutate(date = ymd(as.character(date)))

env_long_raw <- t2$data_env |> 
  select(-c(depth, sample_date, use_maestro, strat_season_2, solar_season, mixing, month)) |> 
  pivot_longer(cols = -c(date, strat_season, sample), names_to = "var", values_to = "value") |> 
  mutate(interpolated = ifelse(is.na(value), TRUE, FALSE))

# AIC: tag interpolated values in the complete dataset for visual flagging
env_long_complete <- env_long_complete |> 
  mutate(interpolated = env_long_raw$interpolated)

env_long_complete |> 
  ggplot(aes(date, value, color = interpolated)) +
  geom_point() +
  facet_wrap(~var, scales = "free")
```

## Check Completeness and Remove NA Variables

```{r}
# AIC: skim gives complete_rate per variable — use this to find columns with any NAs
sum_stats <- skim(t1$data_env)
sum_stats

na_vars <- sum_stats$skim_variable[which(sum_stats$complete_rate < 1)]

t1$data_env <- t1$data_env |> 
  dplyr::select(-any_of(na_vars))

colnames(t1$data_env)

sum_stats <- skim(t1$data_env)
sum_stats

# Remove other variables I don't want
t1$data_env <- t1$data_env |> 
  dplyr::select(-matches(c("depth", "pres", "use_maestro", "date")))
colnames(t1$data_env)
```

## Wilcoxon Tests by Season Group

```{r}
# AIC: testing whether env variables differ significantly across seasonal groupings
t1$cal_diff(group = "strat_season", method = "wilcox")
view(t1$res_diff)

t1$cal_diff(group = "strat_season_2", method = "wilcox")
view(t1$res_diff)

t1$cal_diff(group = "solar_season", method = "wilcox")
view(t1$res_diff)

t1$cal_diff(group = "mixing", method = "wilcox")
view(t1$res_diff)

# Looking at the above tables I am thinking about keep do_sat, cond, temp, pco2, and phycocyanin.
tmp <- list()
for(i in 1:length(colnames(t1$data_env))){
    tmp[[i]] <- t1$plot_diff(measure = colnames(t1$data_env)[i], add_sig_text_size = 5, xtext_size = 12) + theme(plot.margin = unit(c(0.1, 0, 0, 1), "cm"))
}
plot(gridExtra::arrangeGrob(grobs = tmp, ncol = 3))
```

## Pairwise Correlations

```{r}
t1$cal_autocor()
clean_data <- t1$data_env |> 
  select(-matches(c("sample", "strat", "mixing", "solar_season", "month", "turb", "ph_", "do_24")))

plot <- GGally::ggpairs(clean_data)
plot
```

## dbRDA — Community vs. Environment

```{r}
# AIC: clone t1 so we can subset env vars without losing the original object
t2 <- clone(t1)
t2$ data_env <- t2$data_env |> 
  select(-matches(c("sample", "strat", "mixing", "solar_season", "month", "turb", "ph_", "do_24")))
t2$cal_ordination(method = "dbRDA", use_measure = "wei_unifrac")
t2$trans_ordination()
t2$plot_ordination(plot_color = "strat_season")
# the main results of RDA are related with the projection and angles between arrows
# adjust the length of the arrows to show them better
t2$trans_ordination(adjust_arrow_length = TRUE, max_perc_env = 1.5)
# t1$res_rda_trans is the transformed result for plotting
t2$plot_ordination(plot_color = "strat_season", add_sample_label = "sample") +
  ggarrow::geom_arrow_chain(colour = "black") 

t2$cal_autocor()
t2$cal_ordination_anova()
```

## dbRDA by Taxa (Family Level)

```{r}
t2$cal_ordination(method = "RDA", use_measure = "wei_unifrac", taxa_level = "f")
t2$trans_ordination(show_taxa = 50, adjust_arrow_length = TRUE, max_perc_env = 0.5, max_perc_tax = 0.5)
t2$plot_ordination(plot_color = "strat_season", add_sample_label = "sample") +
  ggtitle("RDA of mt_mag at family level")
```

---

# Environmental Analysis — 16S

## Build trans_env Object

```{r}

# First add a "month" column to sample table 
mt_16s$sample_table <- mt_16s$sample_table |> 
  mutate(month = as.factor(month(date)))

# AIC: levels set to water year order (Aug–Jul) rather than calendar order
mt_16s$sample_table$month <- mt_16s$sample_table$month |> 
  factor(levels = c("8", "9", "10", "11", "12", "1", "2", "3", "4", "5", "6", "7"))
mt_16s$sample_table$month

cols <- ncol(mt_16s$sample_table)
t1 <- trans_env$new(dataset = mt_16s, env_cols = c(1:cols), standardize = F, complete_na = T)
```

## Check Completeness and Remove NA Variables

```{r}
sum_stats <- skim(t1$data_env)
sum_stats

na_vars <- sum_stats$skim_variable[which(sum_stats$complete_rate < 1)]

t1$data_env <- t1$data_env |> 
  dplyr::select(-any_of(na_vars))

colnames(t1$data_env)

sum_stats <- skim(t1$data_env)
sum_stats

# Remove other variables I don't want
t1$data_env <- t1$data_env |> 
  dplyr::select(-matches(c("depth", "pres", "use_maestro", "date")))
colnames(t1$data_env)
```

## Wilcoxon Tests by Season Group

```{r}
t1$cal_diff(group = "strat_season", method = "wilcox")
view(t1$res_diff)

t1$cal_diff(group = "strat_season_2", method = "wilcox")
view(t1$res_diff)

t1$cal_diff(group = "solar_season", method = "wilcox")
view(t1$res_diff)

t1$cal_diff(group = "mixing", method = "wilcox")
view(t1$res_diff)

# Looking at the above tables I am thinking about keep do_sat, cond, temp, pco2, and phycocyanin.
tmp <- list()
for(i in 1:length(colnames(t1$data_env))){
    tmp[[i]] <- t1$plot_diff(measure = colnames(t1$data_env)[i], add_sig_text_size = 5, xtext_size = 12) + theme(plot.margin = unit(c(0.1, 0, 0, 1), "cm"))
}
plot(gridExtra::arrangeGrob(grobs = tmp, ncol = 3))

# Calculate correlations across the variables
# t1$cal_autocor()
# clean_data <- t1$data_env |> 
#   select(-matches(c("sample", "strat", "mixing", "solar_season", "month", "turb", "ph_", "do_24")))

# plot <- GGally::ggpairs(clean_data)
# plot
```

## dbRDA — Community vs. Environment

```{r}
t2 <- clone(t1)
t2$ data_env <- t2$data_env |> 
  select(-matches(c("sample", "strat", "mixing", "solar_season", "month", "turb", "ph_", "do_24", "Seq", "fastq")))
t2$cal_ordination(method = "dbRDA", use_measure = "robust.aitchison")
t2$trans_ordination()
t2$plot_ordination(plot_color = "strat_season")
# the main results of RDA are related with the projection and angles between arrows
# adjust the length of the arrows to show them better
t2$trans_ordination(adjust_arrow_length = TRUE, max_perc_env = 2)
# t1$res_rda_trans is the transformed result for plotting
t2$plot_ordination(plot_color = "strat_season", add_sample_label = "sample") +
  ggarrow::geom_arrow_chain(colour = "black") 

t2$cal_autocor()
t2$cal_ordination_anova()
```

## dbRDA by Taxa (Family Level)

```{r}
t2$cal_ordination(method = "RDA", use_measure = "bray", taxa_level = "f")
t2$trans_ordination(show_taxa = 20, adjust_arrow_length = TRUE, max_perc_env = 0.5, max_perc_tax = 0.5)
plot <- t2$plot_ordination(plot_color = "strat_season", add_sample_label = "sample") +
  ggtitle("RDA of mt_16s at family level")
plot
```

---

# Temperature Profile

## Load and Process MATLAB Temperature Data

```{r}
# Import temp data
load("output/data/temp_df_long.RData")
temp_df_long <- temp_df_long |> 
    ungroup()
```

```{r}
library(R.matlab)
library(tidyverse)

# Load metadata-------------------------------------------------------------------------
sw_meta <- read_csv("data/sw_metadata.csv") |>
  mutate(date = mdy(date)) |> 
  dplyr::select(!day:year)

# AIC: strat_season_2 collapses fall into winter for a simpler two-season comparison
sw_meta <- sw_meta |> 
  mutate(strat_season_2 = ifelse(strat_season == "fall", "winter", strat_season))
sw_meta$strat_season <- factor(sw_meta$strat_season, levels = c("summer", "fall", "winter", "spring"), ordered = T)

sw_meta$strat_season_2 <- factor(sw_meta$strat_season_2, levels = c("summer", "winter", "spring"), ordered = T)

# Create another stratification based on the equinox
sw_meta <- sw_meta |> 
  mutate(solar_season = ifelse(date < "2024-12-21", "summer", "winter")) |> 
  mutate(solar_season = ifelse(solar_season == "summer" & date >= "2024-09-22", "fall", solar_season)) |> 
  mutate(solar_season = ifelse(solar_season == "winter" & date >= "2025-03-20", "spring", solar_season)) |> 
  mutate(solar_season = ifelse(date > "2025-06-20", "summer", solar_season))

# Import temp file ----------------------------------------------------------------------------
mat_df <- readMat.default(con = "data/metadata/LSEO24h_temp.mat")

# Extract dataframe we want from 'T' for temp and convert matlab time to poxit time
temp_df <- mat_df$T |> 
  as.data.frame()

# Set colnames and rownames
colnames(temp_df) <- mat_df$dep  
rownames(temp_df) <- mat_df$t 

# AIC: MATLAB serial date (days since Jan 0, 0000) offset to Unix epoch (719529 days)
temp_df <- temp_df |> 
    rownames_to_column(var = "dttm") |>
  mutate(dttm = as_datetime((as.numeric(dttm) - 719529) * 86400)) |> 
    # round to nearest hour to make things easier
  mutate(dttm = round_date(dttm, unit = "hour"))
  #   # Select only columns less than 50m
  # dplyr::select(which(as.numeric(colnames(temp_df)) <= 50))

temp_df_long <- temp_df |> 
  # rownames_to_column(var = "dttm") |>
  # mutate(dttm = as_datetime((as.numeric(dttm) - 719529) * 86400)) |> 
  # # round to nearest hour to make things easier
  # mutate(dttm = round_date(dttm, unit = "hour")) |> 
  pivot_longer(cols = !dttm, names_to = "depth", values_to = "temp") |>
  # Create variable of local time
  mutate(dttm_local = with_tz(dttm, tzone = "US/Central")) |> 
# Add averaged temperature column
  mutate(date = floor_date(dttm_local, unit = "day")) |> 
  group_by(date, depth) |> 
  mutate(temp_c_24h_med = median(temp)) |> 
  ungroup()
```

## Calculate 14-Day Pre-Sample Temperature Average

```{r}
# First create temporary df of sample dates
  samp_dates <- sw_meta |> 
    dplyr::select(sample) |> 
    mutate(sample_date = as_date(sw_meta$date))

# AIC: join_by(closest(...)) assigns each temp obs to the next upcoming sample date
temp_df_long <- temp_df_long |>
  mutate(date = as_date(date)) |>
  left_join(samp_dates, join_by(closest(date <= sample_date))) |>
  mutate(test = as.numeric(sample_date - date)) |> 
  # NEW STEP: Force a strict 14-day window!
  # Keep only the rows where the temp date is within 14 days of the sample date
  # filter(test <= 14) |>
  group_by(sample_date, depth) |> 
  mutate(temp_c_14d_avg = mean(temp, na.rm = TRUE)) |> 
  ungroup()

# OLD VERSION
# locs <- seaprocess::find_near(vec = as_date(temp_df_long$date), vals = as_date(sw_meta$date))
# dates <- as_date(rep(NA, nrow(temp_df_long)))
# for(i in length(locs):1) {
# dates[1:locs[i]] <- as_date(sw_meta$date[i])
# }
# rm(i)
# # Add the 14ish day average
# temp_df_long<-temp_df_long |>
#   mutate(sample_date = dates) |> 
#   group_by(sample_date, depth) |> 
#   mutate(temp_c_14d_avg = mean(temp)) |> 
#   ungroup()

# Create a coefficient of variation to see if we should average or not
temp_df_long <- temp_df_long |> 
  group_by(date) |> 
  mutate(coeff_var = (sd(temp, na.rm = T)/mean(temp, na.rm = T)) * 100) |> 
  ungroup()
```

## Temperature Contour Plot

```{r}

# #Interpolate/grid data first

# install.packages("interp")
# library(interp)
# gridded_data <- temp_df_long |>
#     ungroup() |> 
#     dplyr::select(dttm, depth, temp) |> 
#     drop_na() |> 
#     interp(x = temp_df_long$dttm, 
#                        y = temp_df_long$depth, 
#                        z = temp_df_long$temp)


# AIC: build custom 1°C bin labels to match contour binwidth = 1
temp_breaks <- cut(temp_df_long$temp, 20) |> 
  unique() |> 
  str_remove("\\(") |> 
  str_remove("\\]") |> 
  str_squish() |> 
  str_extract(".*(?=,)") |> 
    as.numeric() |> 
    na.omit() |> 
  sort()

temp_labels_1 <- seq(0, 17, by = 1)
temp_labels_2 <- seq(1, 18, by = 1)

temp_labels <- paste0(temp_labels_1, "-", temp_labels_2)

# Create temp plot
temp_df_long |> 
    ungroup() |> 
    # dplyr::select(dttm, depth, temp) |> 
    drop_na(temp) |> 
    mutate(depth = as.numeric(depth)) |> 
    # filter(depth < 50) |> 
        dplyr::select(date, depth, temp, sample_date) |> 
        # drop_na(temp) |> 
        distinct() |> 
    ggplot(aes(x = date, y = depth, z = temp)) +
      # geom_contour_filled(breaks = breaks) +
      geom_contour_filled(binwidth = 1) +
        # scale_fill_brewer(palette = "RdBu", direction = -1) +
        scale_fill_viridis_d(option = "turbo", labels = temp_labels) +
  labs(x = "Date", y = "Depth (m)", fill = "Temperature (°C)") +
    theme_classic() +
    theme(
      text = element_text(size = 15),
      panel.border = element_rect(colour = "black", fill = "red", linewidth = 3)
    ) +
  scale_y_reverse(expand = c(0,0), n.breaks = 10) +
  scale_x_date(expand = c(0,0), date_breaks = "1 month", date_label = "%b", limits = c(as_date("2024-07-20"), NA), ) +
    guides(fill = guide_legend(ncol = 1, reverse = T)) +
    geom_point(data = samp_dates, aes(x = sample_date, y = 38, z = NULL), color = 'white') +
    # geom_point(data = samp_dates, aes(x = sample_date, y = 22, z = NULL), color = 'white') +
    geom_hline(yintercept = 22, color = 'white', linetype = "dashed" ) +
    # geom_vline(xintercept = sw_meta$date[which(sw_meta$strat_season == "summer")], color = "red") +
    # geom_vline(xintercept = sw_meta$date[which(sw_meta$strat_season == "fall")]) +
    # geom_vline(xintercept = sw_meta$date[which(sw_meta$strat_season == "winter")], , color = "red") +
    # geom_vline(xintercept = sw_meta$date[which(sw_meta$strat_season == "spring")]) +
    theme(text = element_text(size = 25), aspect.ratio = 0.7) +
    NULL

 sw_meta$date[which(sw_meta$strat_season == "summer")]

 quantile(mt_mag$sample_table$temp_c_24h_med)
```

---

# Maestro Sensor Plots

```{r}
load("output/data/maestro_df_clean.RData")
load("output/data/clean_long_m_df.RData")

labels <- read_csv("data/units_cheatsheet.csv")

lookup <- setNames(labels$nice_name, labels$daily_avg)

clean_long_m_df |> 
    filter(variable != "T2", variable != "P") |> 
    ggplot(aes(date, vals, color = variable)) +
    geom_point() +
    facet_wrap(~variable, scales = "free") +
    theme_minimal() +
    labs(color = "Variable", x = "Date", y = NULL)

  noon_midnight_plot <- maestro_df_clean |> 
    dplyr::select(round_date, matches("12h_med")) |> 
    pivot_longer(cols = ! "round_date", names_to = "var", values_to = "vals") |> 
    distinct() |> 
      ggplot(aes(round_date, vals, color = var)) +
    geom_point() +
    facet_wrap(~var, scales = "free") +
    theme_minimal() +
    labs(color = "Variable", x = "Date", y = NULL) +
    theme_classic()
noon_midnight_plot

daily_plot <- maestro_df_clean |> 
    dplyr::select(date, matches("24h_avg"), -matches("mtemp_2_")) |> 
    pivot_longer(cols = !date, names_to = "var", values_to = "vals") |> 
    distinct() |>
      ggplot(aes(date, vals, color = var)) +
    geom_line() +
    facet_wrap(~var, scales = "free", labeller = as_labeller(lookup) ) +
    theme_minimal() +
    labs(color = "Variable", x = "Date", y = NULL) + 
    theme_classic() +
    theme(legend.position = "none")

daily_plot

# Temp vs. DO
  plot <- maestro_df_clean |> 
    ggplot(aes(x = T, y = DO, color = date)) +
    geom_line() +
    scale_color_viridis_c(trans = "date", direction = 1) +
    theme_classic() +
    labs(x = "Temp (C)", y = "DO (umol/kg)")
    plot
  plotly::plotly_build(plot)

# Temp vs. DO saturation
    plot <- maestro_df_clean |> 
    ggplot(aes(x = T, y = do_sat_percent, color = date)) +
    geom_point() +
    scale_color_viridis_c(trans = "date", direction = 1) +
    theme_classic() +
    labs(x = "Temp (C)")
    plot
  plotly::plotly_build(plot)

# Temp vs. pH
    plot <- maestro_df_clean |> 
    ggplot(aes(x = T, y = pH, color = date)) +
    geom_point() +
    scale_color_viridis_c(trans = "date", direction = 1) +
    theme_classic() +
    labs(x = "Temp (C)", y = "pH")
    plot
  plotly::plotly_build(plot)

# Temp vs. Chlorophyll
    plot <- maestro_df_clean |> 
    ggplot(aes(x = T, y = CHL, color = date)) +
    geom_point() +
    scale_color_viridis_c(trans = "date", direction = 1) +
    theme_classic() +
    labs(x = "Temp (C)", y = "CHL")
    plot
  plotly::plotly_build(plot)

  colnames(maestro_df_clean)
```

## Sample Table Variables Over Time

WM03 where things look all sorts of weird has high chla and cdom and do saturation. The maestro temp is pretty high but it seems like the rest is pretty standard.

```{r}
mt_mag$sample_table |> 
  view()

mt_mag$sample_table |> 
  select(sample, date, contains("24h")) |>
  mutate(is_WM03 = sample == "WM03") |> 
  pivot_longer(cols = -c(sample, date, is_WM03), names_to = "var", values_to = "val") |> 
  ggplot(aes(date, val, group = var, color = is_WM03)) +
  geom_point() +
  geom_line() +
  facet_wrap(~var, scales = "free") +
  theme_classic()
```

---

# Storm Investigation

I want to look and see if there was a storm around WM03 sampling time.

## WM Weather Station

```{r}
  wm <- read_table("data/metadata/wm_wx_data_2024.txt") |> 
  rbind(read_table("data/metadata/wm_wx_data_2025.txt"))

units <- data.frame(col = colnames(wm), unit = as_vector(wm[1,])) |> 
  rownames_to_column("delete") |> 
  select(-delete)

clean_wm <- wm |> 
  mutate(across(everything(), as.numeric)) |> 
  filter_out(if_all(everything(), is.na)) |> 
  mutate(dttm = ymd_hm(paste(`#YY`, MM, DD, hh, mm))) |> 
  select(-(`#YY`:mm)) |> 
  rename_with(~str_to_lower(.x)) |> 
  select(-tide, -vis, -mwd, -apd, -dpd) |> 
  mutate(across(!matches("dttm"), ~as.numeric(.x))) |> 
  # AIC: wspd > 80 likely instrument errors; set to NA
  mutate(wspd = ifelse(wspd > 80, NA, wspd),
         date = round_date(dttm, unit = "day")) |> 
  group_by(date) |> 
  mutate(across(wdir:dewp, ~mean(.x), .names = "mean_{.col}"), 
         across(wdir:dewp, ~max(.x), .names = "max_{.col}"), 
         across(wdir:dewp, ~median(.x), .names = "med_{.col}"))|> 
  ungroup()

# Plot wind speed over dttm
clean_wm|> 
  ggplot(aes(date, mean_wspd)) +
  geom_point() +
  geom_vline(data = mt_mag$sample_table, aes(xintercept = date, color = sample == "WM03"))

clean_wm |> 
  filter(date >= "2024-07-01" & date <= "2025-01-01") |> 
  pivot_longer(cols = -c(date, dttm), names_to = "var", values_to = "val") |> 
  ggplot(aes(date, val)) +
  geom_point() + 
  facet_wrap(~var, scales = "free") +
  geom_vline(data = mt_mag$sample_table, aes(xintercept = date, color = sample == "WM03"))
```

Hmm, okay from this data maybe it is a wind direction thing? The wind speed looks minimal compared to the rest of the surrounding dates.

## ROAM4 Weather Station

```{r}
  rock <- read_table("data/metadata/roam4_wx_data_2024.txt") |> 
  rbind(read_table("data/metadata/roam4_wx_data_2025.txt"))

units <- data.frame(col = colnames(rock), unit = as_vector(rock[1,])) |> 
  rownames_to_column("delete") |> 
  select(-delete)

clean_rock <- rock |> 
  mutate(across(everything(), as.numeric)) |> 
  filter_out(if_all(everything(), is.na)) |> 
  mutate(dttm = ymd_hm(paste(`#YY`, MM, DD, hh, mm))) |> 
  select(-(`#YY`:mm)) |> 
  rename_with(~str_to_lower(.x)) |>
  select(-contains("g")) |> 
  mutate(across(!matches("dttm"), ~as.numeric(.x))) |> 
  mutate(wspd = ifelse(wspd > 80, NA, wspd),
         date = as_date(round_date(dttm, unit = "day"))) |> 
  group_by(date) |> 
  mutate(across(is.numeric, ~mean(.x), .names = "mean_{.col}"), 
         across(is.numeric, ~max(.x), .names = "max_{.col}"))|> 
  ungroup()

# Plot wind speed over dttm
clean_rock|> 
  filter(date >= "2024-07-01" & date <= "2025-06-01") |> 
  pivot_longer(cols = -c(date, dttm), names_to = "var", values_to = "val") |> 
  ggplot(aes(date, val)) +
  geom_point() + 
  facet_wrap(~var, scales = "free") +
  geom_vline(data = mt_mag$sample_table, aes(xintercept = date, color = sample == "WM03"))
```
