---
title: "Initial Explorations"
format: html
editor: visual
---

# Load required packages

```{r}
library(tidyverse) 
library(vegan) 
library(Polychrome) 
library(forcats)
library(paletteer)
```

# Read in the data

```{r}

b_df <- readr::read_tsv("Sylph_TaxAbund_out.tsv")
```

# Clean data

```{r}

# Create vectors of column names
clade_cols <- stringr::str_extract_all(b_df$clade_name[8], "[a-z]{1}(?=_)")

clade_cols <- as.vector(clade_cols[[1]]) %>% print()

site_cols <- 
  stringr::str_extract(colnames(b_df), "(?<=/).{4}") %>%   
  na.omit() %>% 
  print() 

colnames(b_df)[2:ncol(b_df)] <- site_cols

#Spread columns out, clean names

b_df <- b_df %>% tidyr::separate_wider_delim(cols = clade_name, delim = "|", names = clade_cols, too_few = "align_start") %>% 
  # Drop all rows that don't get down to a taxa 
  tidyr::drop_na("t") %>%
  #remove additional characters and make data columns numeric
  mutate(across(clade_cols, ~ str_remove(., "[a-z]__")))

# Quick check to see if abundances are all still 100 
colSums(b_df[9:ncol(b_df)])

# Load metadata
sw_meta <- read_csv("sw_metadata.csv")

#Create long df 
long_b_df <- b_df %>% pivot_longer(cols = all_of(site_cols), names_to = "sample", values_to = "abundance")

long_b_df <- long_b_df %>% 
  mutate(date = sw_meta$date[match(long_b_df$sample, sw_meta$Sample)]
  )
long_b_df$date <- mdy(long_b_df$date)



shades$country_name <- country$name[match(shades$group, country$code)]

```

# Plots

## Bar chart

```{r}
#First create a unique palette based on how many categories there are 
palette <- length(unique(b_df$p)) %>% glasbey.colors() %>% unname()

long_b_df$p <- fct_reorder(long_b_df$p, long_b_df$abundance)

# Plot
long_b_df %>% ggplot(aes(x = sample, y = abundance, fill = p)) + geom_bar(stat = "identity") + 
  # scale_fill_discrete() +
  scale_fill_manual(values = palette)+
  scale_x_discrete(guide = guide_axis(n.dodge = 2), 
                   labels = as.character(format(long_b_df$date, "%m-%d"))
                   )+
  labs(x = "Date", y = "Abundance", legend = "Phyla", fill = "Phyla") +
  theme_classic() +
  NULL

long_b_df %>% ggplot(aes(x = sample, y = abundance, fill = p)) + geom_bar(stat = "identity") + 
  # scale_fill_discrete() +
  scale_fill_manual(values = palette)+
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  labs(x = "Sample", y = "Abundance", legend = "Phyla", fill = "Phyla") + 
  theme_classic() +
  NULL
```

## Interpretation

**Major Phyla:** 1. Pseudomonadota 2. Bacteriodita 3. Actinomycetota 4. Planctomycetota 5. Verrucomicrobiota

**Biggest changes:** 1. Really big drop in Bacteroidota between WM02 and WM03. 2. Pseudomonadota abundance grows throughout the year 3. Actinomycetota seem to drop off in the middle samples. 4. Planctomycetota are more significant at the start but then drop off by \~WM16 5. A small uptick in Thermoproteota in the middle months

## Class chart

Only of 5 major families

```{r}
# Plot

long_b_df %>% 
  filter(p == c("Pseudomonadota", "Bacteriodita", "Actinomycetota", "Planctomycetota", "Verrucomicrobiota"), abundance > 0.1) %>%
           ggplot(aes(x = md(date), y = abundance, fill = fct_reorder(f, abundance)))+ 
  geom_bar(stat = "identity", position = position_fill()) + 
  # scale_fill_discrete() +
  # scale_fill_manual(values = palette)+
  scale_color_paletteer_d() +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  NULL
```

## NMDS

```{r}
# NMDS model

nmds <- b_df %>% 
  select(all_of(site_cols)) %>% 
  t() %>% 
  # decostand("hellinger") %>% 
  metaMDS(distance="bray", autotransform=FALSE, binary=FALSE, noshare=TRUE, zerodist="ignore")

# Extract scores

nmds_scores <- scores(nmds, display = "sites") %>% 
  data.frame() %>%
  rownames_to_column(var = "sample")

# Basic NMDS plot 
nmds_basic_plot <- 
  ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, color = sample)) + 
  geom_point(size = 3, alpha = 0.8) + 
  labs(title = "NMDS Ordination of WM Data", subtitle = paste("Stress =",      round(nmds$stress, 3)), x = "NMDS1", y = "NMDS2", color = "Sample") +
  scale_color_viridis_d() +
  theme_classic()

nmds_basic_plot

# Run stressplot 
stressplot(nmds)
```

## Interpretation

Looks like there are about three distinct groups which follow a yearly cycle pretty well. **Next Steps:** See how this maps on to physical changes, possibly isothermal periods.

# To Do:

1.  Get metadata from Jay
2.  Ask Cody: are the taxa we are seeing whole water columns? Just benthic?